---
title: 'census 2020: some quick visuals of demographic change'
author: ''
date: '2021-08-13'
slug: census-2020-some-quick-visuals
categories: []
tags:
  - rstats
  - census
  - geo
twitterImg: img/census2020.png
banner: banners/census2020.png
description: 'Ealry access to 2020 Census data via the PL94171 package'
---

## Intro 

> A quick/simple post using the [PL94171](https://github.com/CoryMcCartan/PL94171) package to access Census 2020 counts.  While census data will not be API-accessible until October, redistricting files are presently available -- the `PL94171` package facilitates easy access to these files. Some quick visualizations of demographic change in the state of New Mexico.


## Get data

Using several of the functions included in the `PL94171` package, we build a simple wrapper below to extract census counts for multiple census years.  

```{r message=FALSE, warning=FALSE}
library(tidyverse)

get_pl_data <- function(x, state, level){

  y <- PL94171::pl_read(PL94171::pl_url(state, x))
  pl <- PL94171::pl_subset(y, sumlev = level)
  PL94171::pl_select_standard(pl, clean_names = TRUE)
}
```


The we apply the function, collecting county-level data for the state of New Mexico for the last three decennial censuses.

```{r eval=FALSE}
yrs <- c(2000, 2010, 2020)

x0 <- lapply(yrs, get_pl_data, state = 'NM', level = '050')
names(x0) <- yrs
x1 <- data.table::rbindlist(x0, idcol = 'year')
```

```{r include=FALSE}
#saveRDS(x1, 'nm-census-counts.rds')
x1 <- readRDS('/home/jtimm/pCloudDrive/GitHub/nm-census-counts.rds')
```



## County level change by sub-group: 2000-2020

Below we re-structure the data some, and download New Mexico county information via the `tigris` package.

```{r message=FALSE, warning=FALSE}
x2 <- x1 %>%
  select(year, GEOID, pop:pop_two) %>%
  gather(key = 'race', value = 'count', -year:-pop) %>%
  mutate(prop = count/pop)
```


```{r}
head(x2) %>% knitr::kable()
```


```{r include=FALSE}
nm <- tigris::counties(state = 'NM', cb = T)
```

```{r eval=FALSE}
nm <- tigris::counties(state = 'NM', cb = T)
```


A homemade palette:

```{r}
gen_pal <- c('#ead8c3', 
             '#eeeeee',
             '#437193',
             '#7da6aa',
             '#b0bcc1',
             '#55752f', 
             '#dae2ba',
             '#eb7f6b'
             )
```


Population change by sub-group for a sample of counties in New Mexico:

```{r fig.height=8, message=FALSE, warning=FALSE}
set.seed(999)
x2 %>%
  inner_join(nm %>% sample_n(12)) %>%
  
  ggplot(aes(x = as.integer(year), 
             y = prop, 
             fill = race)) +
  geom_area(stat = "identity",
            color = 'white',
            alpha = 0.85) +
  geom_hline(yintercept = .5, 
             linetype = 4,
             color = 'white') +
  scale_fill_manual(values = gen_pal) +
  scale_x_continuous(breaks=seq(2000,2020,10)) +
  xlab('') + ylab('') +
  theme_minimal() +
  theme(legend.position="bottom",
        legend.title = element_blank(),
        axis.text.x = element_text(angle = 90, hjust = 1)) +
  facet_wrap(~NAME, ncol = 4) +
  ggtitle('Population composition by county: 2000-2020')
```


## A quick map

```{r}
x3 <- x2 %>%
  filter(race == 'pop_hisp' &year != 2000) %>%
  select(year, GEOID, prop) %>%
  spread(year, prop) %>%
  mutate(delta = round(`2020` - `2010`, 3))

head(x3) %>% knitr::kable()
```


The map below details percent change for the Hispanic population in New Mexico from 2010 to 2020. 

```{r fig.height=6, message=FALSE, warning=FALSE}
nm %>%
  left_join(x3) %>%
  ggplot() + 
  geom_sf(aes(fill = delta),
          color = 'darkgray',
          alpha = .85,
          lwd = .2) +

  scale_fill_distiller(palette = "BrBG",  
                        limit = max(abs(x3$delta)) * c(-1, 1)) +

  theme_minimal() + 
  theme(axis.title.x=element_blank(), 
          axis.text.x=element_blank(),
          axis.title.y=element_blank(),
          axis.text.y=element_blank(),
          legend.title=element_blank(),
          legend.position = 'bottom', 
          complete = F) +
  labs(title = 'Percent Change Hispanic Population in New Mexico', 
       subtitle = 'by County: 2010 to 2020')
```


## Census 2020
